# Loading Fisheries Data

```{r include=FALSE}
knitr::opts_chunk$set(warning = F, message = F, error = F)
```

Next, you'll load the data you'll be using for the analysis using the `read_csv` function. First, create a new folder in your working directory called "_data". Next, download the following two files onto your computer. Right click on each link, and save it into your "_data" folder.

[Right click and save-link-as to download landings data](https://raw.githubusercontent.com/SFG-UCSB/fishery-manageR/master/_data/sample_landings_data.csv)

[Right click and save-link-as to download life history data](https://raw.githubusercontent.com/SFG-UCSB/fishery-manageR/master/_data/life_history_parameters.csv)

Next, you'll load these two csv files into R. First load the landings data which includes catch, effort, and length measurements and store these data to a data frame called `landings_data`. Next, load the life history parameter data file and store these data to a data frame called `life_history_parameters`.

```{r}
landings_data <- read_csv("_data/sample_landings_data.csv")

life_history_parameters <- read_csv("_data/life_history_parameters.csv")
```

Let's take a quick look at what's in each of these data frames. We can use the `head` function to get a quick summary.

```{r}
head(landings_data)
```

The `landings_data` data frame is from a fishery-dependent landing site survey. The species included in this data set is *Caesio cuning*, a yellowtail fusilier. You can see the data frame has eight columns, with each row being an individual fish catch measurement. The columns include the year and date when the measurement was collected, the fishing trip ID, how many hours were fished for each trip, what gear was used, the species, the length of the fish, and the weight of the fish. We'll be able to use these data to create length-frequency histograms that describe the size structure of the population, as well as trends in catch and CPUE. 

One thing we should do is format the year and date columns to ensure that R knows these represent years and dates as we humans think of them. We can use the `mdy` and `year` functions from the lubridate package to do this.

```{r}
## Start with the landings data frame
landings_data <- landings_data %>% 
  ## Turn the date column into a date format that R recognizes
  mutate(Date = mdy(Date)) %>% 
  ## Create a year column using a year format R recognizes
  mutate(Year = year(Date))
```

Let's next look at the `life_history_parameters` data frame.

```{r}
head(life_history_parameters)
```
The `life_history_parameters` data frame includes several important life history parameters that were found doing a literature review of this species. *L_infinity*, *k*, and *t0* describe the age-to-length relationship using the von Bertalanffy growth equation. *M* represents natural mortality. *Wa* and *Wb* describe the length-to-weight relationship. *m50* and *m95* describe the lengths at which 50% and 95% of the fish are mature, respecitvely. *maxAge* is the maximum age of the fish.